---
title: "Add OrthoFinder Gene Symbols"
output: html_notebook
---

This R notebook takes the output from OrthoFinder, converts the protein-protein relationships to gene-gene relationships, and adds them to the ortholog table.

```{r}
library(plyr)
library(tidyverse)
```

## Format OrthoFinder results

First, read in OrthoFinder results:

```{r}
orthofinder_pairs <- read.table("./orthofinder/Results_date_of_run/Orthologues/Orthologues_target_proteins/target_proteins__v__reference_proteins.tsv",
                                header = TRUE,
                                sep = "\t")
colnames(orthofinder_pairs)[2:3] <- c("mikado_id","orthofinder_gene")
head(orthofinder_pairs)
```

Remove Orthogroup column, as it is not needed

```{r}
orthofinder_pairs$Orthogroup <- NULL
```

In each column, you will see many protein or RNA feature names grouped together by commas in each row (if you followed the tutorial up to this point, you are likely looking at human RNA feature names that are formatted like "rna-id" in the right column). We need to expand by those values to get one protein name in each column.

```{r}
orthofinder_pairs <- tidyr::separate_longer_delim(orthofinder_pairs, 
                                   "mikado_id", 
                                   delim = ", ")
orthofinder_pairs <- tidyr::separate_longer_delim(orthofinder_pairs, 
                                   "orthofinder_gene", 
                                   delim = ", ")
```

Remove decimals from "mikado_id" column as they represent splice variants and come from the same gene and remove duplicate rows

```{r}
orthofinder_pairs$mikado_id <- gsub("\\.[0-9]+$","",
                                    orthofinder_pairs$mikado_id)
orthofinder_pairs <- dplyr::distinct(orthofinder_pairs)
```

## Gene name conversion

Read in reference GFF file (in this example we're using the human GFF)

```{r}
reference <- as.data.frame(rtracklayer::readGFF("reference.gff"))
```

Isolate the appropriate columns representing the "protein name" used by OrthoFinder (e.g. "ID") and the desired gene symbol column (e.g. "gene"). Remove NAs and duplicates from this data frame. Rename the columns so that the "protein name" column matches "orthofinder_gene" so that we can eventually merge the "reference" data frame with the "orthofinder_pairs" data frame, using that column to join the rows.


```{r}
reference <- reference[,which(colnames(reference) %in% c("ID","gene"))]
reference <- reference[!is.na(reference$gene),]
reference <- dplyr::distinct(reference)
colnames(reference) <- c("orthofinder_gene", "gene")
head(reference)
```

Use `left_join` from `dplyr` to add the "reference" data frame to the "orthofinder_pairs" data frame

```{r}
orthofinder_pairs <- dplyr::left_join(orthofinder_pairs, reference, by = "orthofinder_gene")
head(orthofinder_pairs)
```

We can now replace the "orthofinder_gene" column with "gene", and then remove the "gene" column. After this, we will also remove redundant rows since different transcript names may share the same gene name.

```{r}
orthofinder_pairs$orthofinder_gene <- orthofinder_pairs$gene
orthofinder_pairs$gene <- NULL
orthofinder_pairs <- dplyr::distinct(orthofinder_pairs)
head(orthofinder_pairs)
```

Same as with the LiftOff and TOGA results, we will combine multiple gene symbols that represent the same Mikado ID with semi-colons, and add copy numbers to identical gene symbols that represent different Mikado IDs.

Start by sorting:

```{r}
orthofinder_pairs <- orthofinder_pairs[order(orthofinder_pairs$mikado_id,
                         orthofinder_pairs$orthofinder_gene),]
```

Now collapse by Mikado ID and add the semi-colons if necessary:

```{r}
orthofinder_pairs <- plyr::ddply(orthofinder_pairs,
                       "mikado_id",
                       summarize,
                       orthofinder_gene = paste(orthofinder_gene, collapse = ";"))
head(orthofinder_pairs)
```

Add copy numbers:

```{r}
orthofinder_pairs$orthofinder_gene <- make.unique(orthofinder_pairs$orthofinder_gene, sep = "-copy")
```

## Add to gene symbol table

Read in the gene symbol tables that you made earlier

```{r}
mikado_df <- read.table("gene_symbols.tsv", sep = "\t", header = TRUE)
mikado_df_noCopies <- read.table("gene_symbols_noCopies.tsv", sep = "\t", header = TRUE)
```

Use `left_join` to add the new gene symbols to both tables, removing the copies from the "no copies" table

```{r}
mikado_df <- dplyr::left_join(mikado_df,
                       orthofinder_pairs,
                       by = "mikado_id")
mikado_df_noCopies <- dplyr::left_join(mikado_df_noCopies,
                       orthofinder_pairs,
                       by = "mikado_id")
mikado_df_noCopies$orthofinder_gene <- gsub("-copy.*", "", mikado_df_noCopies$orthofinder_gene)
```

Save both tables

```{r}
write.table(mikado_df, file = "gene_symbols.tsv",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = TRUE)

write.table(mikado_df_noCopies, file = "gene_symbols_noCopies.tsv",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = TRUE)
```





